---
title: "Coding Camp Final Project"
author: "Zander Meitus"
date: "9/8/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
For many beaches along and near the Chicago lakefront, the University of Illinois at Chicago conducts daily water quality testing from around Memorial Day until about Labor Day looking at levels of a bacteria called Enterococci. The data is posted online by the Chicago Park System. Colored flags are also raised at each beach depending on the most recent reading: (bacteria levels < 1000 CCE)
- green: water quality is safe for swimming  
- yellow: water quality is unsafe for swimming  (bacteria levels > 1000 CCE)

Water quality can vary greatly across the beaches, and some beaches are cleaner than others.  

``` {r libraries & data, echo = F}
setwd('~/Documents/coding-camp/final-project/')


# READ LIBRARIES & DATA ---------------------------------------------------

library(tidyverse)
library(stringr)
library(lubridate)
library(viridis)
library(emo)
library(corrplot)

# source: https://data.cityofchicago.org/Parks-Recreation/Beach-Lab-Data/2ivx-z93u
beach_lab_data <- read_delim('Beach_Lab_Data.csv', delim = ';')
```

```{r clean data, echo  = FALSE }
```

```{r analyze data }

```

